---
title: "LAGOS Spatial Analysis"
author: "Casey Menick"
date: "02/21/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---



```{r setup, include=FALSE}
library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
```


# LAGOS Analysis


## Loading in data


### First download and then specifically grab the locus (or site lat longs)

```{r data-read}
# #Lagos download script
LAGOSNE::lagosne_get(dest_folder = LAGOSNE:::lagos_path())


#Load in lagos
lagos <- lagosne_load()

#Grab the lake centroid info
lake_centers <- lagos$locus
```



### Convert to spatial data
```{r}
#Look at the column names
#names(lake_centers)

#Look at the structure
str(lake_centers)

#View the full dataset
#View(lake_centers %>% slice(1:100))

spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326) %>%
  st_transform(2163)

#Subset for plotting
subset_spatial <- spatial_lakes %>%
  slice(1:100) 

subset_baser <- spatial_lakes[1:100,]

#Dynamic mapviewer
mapview(subset_spatial)

```


### Subset to only Minnesota

```{r}
states <- us_states()

#Plot all the states to check if they loaded
#mapview(states)
minnesota <- states %>%
  filter(name == 'Minnesota') %>%
  st_transform(2163)

#Subset lakes based on spatial position
minnesota_lakes <- spatial_lakes[minnesota,]

#Plotting the first 1000 lakes
minnesota_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha')
```



# In-Class work


## 1) Show a map outline of Iowa and Illinois (similar to Minnesota map upstream)

```{r}

ILIA <- states %>%
  filter(name == 'Illinois' | name == 'Iowa') %>%
  st_transform(2163)

# view outline of Illinois and Iowa
mapview(ILIA)

#Subset lakes based on spatial position
ILIA_lakes <- spatial_lakes[ILIA,] %>% 
    mutate(lake_area_logha = log(lake_area_ha))

#Plotting the first 1000 lakes
ILIA_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha')

```



## 2) Subset LAGOS data to these sites, how many sites are in Illinois and Iowa combined? How does this compare to Minnesota?

Minnesota has many more lakes (29,038) than Illinois and Iowa combined (16,466). That's more than the land of 10,000 lakes!

```{r}

length(unique(minnesota_lakes$lagoslakeid))
length(unique(ILIA_lakes$lagoslakeid))
```



## 3) What is the distribution of lake size in Iowa vs. Minnesota?


```{r}
#create iowa lakes file
iowa <- states %>%
  filter(name == 'Iowa') %>%
  st_transform(2163)

iowa_lakes <- spatial_lakes[iowa,]


#combine ia and mn data, add statename to lake
MNIA_lakes <- rbind(iowa_lakes %>% mutate(state="Iowa"),minnesota_lakes %>% mutate(state="Minnesota")) %>% 
  mutate(lake_area_logha = log(lake_area_ha))

ggplot(MNIA_lakes,aes(lake_area_logha)) +
  geom_histogram(bins=25) +
  facet_wrap(~state)

```


## 4) Make an interactive plot of lakes in Iowa and Illinois and color them by lake area in hectares

```{r}
#Plotting the first 1000 lakes
ILIA_lakes %>%
  arrange(-lake_area_logha) %>%
  mapview(.,zcol = 'lake_area_logha')
```


## 5) What other data sources might we use to understand how reservoirs and natural lakes vary in size in these three states? 

This analysis has potential to be improved by incorporating other data!

Other data sources could include state records of lakes, satellite imagery of lake boundaries over time, property records, or other research efforts.

It would also be interesting to see how sizes of waterbodies vary when comparing reservoirs vs. natural lakes, and what percent of lakes are natural between the three states. I'd be curious to see if the definition of a "lake" varies between any of these states as well. I'd also like to know how these numbers have changed over time, either in the creation/destruction of lakes, or natural changes to lake boundaries.
