---
title: "Lake Water Quality Analysis"
author: "Matthew Ross"
date: "02/23/2021"
output: html_document
---



```{r setup, include=FALSE}
library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
#mapviewOptions(fgb=FALSE)
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
library(lubridate) #For dealing with date and time
library(kableExtra)
```


# LAGOS Analysis


## Loading in data


### First download and then specifically grab the locus (or site lat longs)
```{r data-read, warnings = FALSE}
#Lagos download script
#lagosne_get(dest_folder = LAGOSNE:::lagos_path(),overwrite=T)

#Load in lagos
lagos <- lagosne_load()


#Grab the lake centroid info
lake_centers <- lagos$locus

# Make an sf object 
spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326)

#Grab the water quality data
nutr <- lagos$epi_nutr

#Look at column names
#names(nutr)
```

### Subset columns nutr to only keep key info that we want


```{r}
clarity_only <- nutr %>%
  select(lagoslakeid,sampledate,chla,doc,secchi) %>%
  mutate(sampledate = as.character(sampledate) %>% ymd(.))

```


### Keep sites with at least 200 observations 

```{r}

#Look at the number of rows of dataset
#nrow(clarity_only)

chla_secchi <- clarity_only %>%
  filter(!is.na(chla),
         !is.na(secchi))

# How many observatiosn did we lose?
# nrow(clarity_only) - nrow(chla_secchi)


# Keep only the lakes with at least 200 observations of secchi and chla
chla_secchi_200 <- chla_secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n()) %>%
  filter(count > 200)


```


### Join water quality data to spatial data

```{r}
spatial_200 <- inner_join(spatial_lakes,chla_secchi_200 %>%
                            distinct(lagoslakeid,.keep_all=T),
                          by='lagoslakeid')


```

### Mean Chl_a map

```{r}
### Take the mean chl_a and secchi by lake

mean_values_200 <- chla_secchi_200 %>%
  # Take summary by lake id
  group_by(lagoslakeid) %>%
  # take mean chl_a per lake id
  summarize(mean_chl = mean(chla,na.rm=T),
            mean_secchi=mean(secchi,na.rm=T)) %>%
  #Get rid of NAs
  filter(!is.na(mean_chl),
         !is.na(mean_secchi)) %>%
  # Take the log base 10 of the mean_chl
  mutate(log10_mean_chl = log10(mean_chl))

#Join datasets
mean_spatial <- inner_join(spatial_lakes,mean_values_200,
                          by='lagoslakeid') 

#Make a map
mapview(mean_spatial,zcol='log10_mean_chl')

```


# Class work

## 1) What is the correlation between Secchi Disk Depth and Chlorophyll a for sites with at least 200 observations?

Secchi disk depth is negatively exponentially correlation with secchi disk depth. As chlorophyll increases, there becomes a point where very little light is able to penetrate the water beyond a short distance. At a certain point, it does not matter how much more additional chlorophyll is in the water, as the light cannot penetrate anyway.

```{r}
#Your code here
ggplot(chla_secchi %>% 
         group_by(lagoslakeid) %>% 
         summarise(meanchla=mean(chla),
                   meansecchi=mean(secchi)),aes(meanchla,meansecchi)) + 
  geom_point()
```


## 2) What states have the most data? 

```{r, warnings = FALSE}
# get count for each lake id
lake_centers <- lake_centers %>%
  group_by(lagoslakeid,nhd_long,nhd_lat,state_zoneid) %>%
  summarise(n=n())

# join data to include state names

lake_states <- lagos$state
lake_statecenters<-left_join(lake_centers,lake_states,"state_zoneid")

# group by state and summarise to find total count
lake_obsn <- lake_statecenters %>% 
  group_by(state_name) %>% 
  summarise(n=sum(n)) %>% 
  arrange(desc(n)) %>% 
  drop_na()

states <- us_states() %>% 
  mutate(state_name=name)

# print table of counts
kable(lake_obsn)

# make map of counts by state
lake_statecenterboundaries <- left_join(lake_obsn,states,"state_name") %>% 
  drop_na()

lake_countmap <- st_as_sf(lake_statecenterboundaries)

mapview(lake_countmap,zcol='n')
```

## 3) Is there a spatial pattern in Secchi disk depth for lakes with at least 200 observations?

Lakes that were further from urban areas tended to have higher secchi depth disks, indicating clearer water. This could be due to nutrient runoff from high population areas leading to high amounts of chla, and lower visibility in the water.

```{r}

spatial_200 <- st_as_sf(left_join(chla_secchi_200,lake_centers,"lagoslakeid"),coords=c("nhd_long","nhd_lat"))

mapview(spatial_200,zcol='secchi')

```


